Augmentation of a Historical Harpsichord Keyboard Replica for Haptic-Enabled Interaction in Museum Exhibitions

### 2025-01-21 src/article.tex

% With the advent of digital musical instruments in the 20th century there has
% been a decoupling between interface and sound generation when playing.
% Consequently, aspects of the sensory experience of playing have been lost,
% namely the sensation of vibration and mechanical impedance that a traditional
% musical instrument provides.

% The project aims to restore the tactile and haptic response of historical
% harpsichords for an exhibition at the Museo di San
% Colombano\footnote{https://genusbononiae.it/en/san-colombano-tagliavini-collection/}
% in Bologna. San Colombano is home to the Tagliavini Collection, a collection of
% approximately 70 historical keyboard instruments in working order
% \cite{Tagliavini2007} gifted to the museum by Luigi Ferdinando
% Tagliavini in 2010 \cite{SanColombano2010,Carlino2010}.

% Some items in the Tagliavini collection are no longer in playing condition or
% are too fragile to continue to be available to museum visitors. One of those
% fragile instruments is a 1547 Harpsichord in the Italian style by Alessandro
% Trasuntino \cite{Wraight2024}\footnote{examples of similar instruments by
% Trasuntino available online in the Royal College of Music Collection (1531:
% \url{https://museumcollections.rcm.ac.uk/collection/Details/collect/58}) and the
% Musée de la musique collection (1538:
% \url{https://carmentis.kmkg-mrah.be/eMP/eMuseumPlus?service=ExternalInterface&objectId=106088})
% }, which was used as the basis for an interactive exhibition arranged by San
% Colombano. Part of the exhibit is a keyboard that replicates the mechanism of
% the Trasuntino that aims to provide a similar tactile playing experience to that
% of the original.

% The project was also open-sourced to allow for others to easily recreate and
% iterate on the designs.

% The "Haptic Key" project by Timmermans et al. \cite{Timmermans2020} began with
% the model of a piano key. A 3-key cross-sectional model of a harpsichord
% mechanism \ref{fig:3key} was provided to the NEMUS project by San Colombano as a
% test bed for sensors and as proof of concept for a larger scale interface. Each
% key of the model has 2 sets of strings that are under tension but not at any
% specific musical pitch. The strings are there to provide the same resistance to
% the jack plectra as would be found on an instrument. The mechanism is in the
% Italian style of the XVI-XIX century with 2 jacks per key and their
% corresponding strings tuned in unison \footnote{``8'-8' È la disposizione normale
% dei clavicembali italiani'' \cite{Tagliavini1987}}.

% A suitable sensing system was found to reliably detect when a jack had plucked a
% string independently and then send a corresponding MIDI message to a connected
% audio engine. A full scale 49-key scale model was commissioned with and then
% built by the luthier \anon{Roberto Livi} in the same style as the 1547 Trasuntino
% \ref{fig:teaser}. The sensor system was installed into the full-scale model and
% then set up for exhibition at San Colombano. Visitors are invited to play on the
% interface, which currently triggers samples on a connected sound engine. The
% paper looks at the design decisions, processes, and considerations that need to
% be made when undertaking a project of this kind.

% Overview of fabrication methods, techniques that can be employed.
% Finally a discussion of the limitations and areas in which the core
% project can be improved and developed further.

% In this context, a `traditional' instrument is one whose interface for
% generating sound and the sound generating mechanism are fundamentally
% coupled. A piano requires a key press which triggers a hammer, which
% strikes a string which vibrates a body and air cavity. This system is
% tied together and cannot be separated. A synthesizer, be it a theremin,
% a control voltage synthesizer, a MIDI controller, these are all
% instruments whose interface are decoupled from their sound generation.
% This decoupling presents a problem that then needs to be addressed. How
% do we re-introduce this dimension to the experience of playing a musical
% instrument? How do we, and is it possible to, parameterise and preserve
% this experience? This is central problem that is trying to be solved in
% \cite{Nichols2002, Timmermans2020, McAlpine2014, Baldwin2016}.

% This is particularly pertinent as it relates to the conservation of musical
% instruments, as they represent both historic objects, but also tools for
% creating music. Historical musical instruments present an interesting
% problem for cultural heritage as they are both a physical object, but
% also a tool for making music. This begs the question, can we use haptics
% to preserve experiences of the past? The proposed project intends to
% develop further the methodologies implemented in musical haptics
% projects \cite{MusicalHaptics2018} such as \cite{Timmermans2020}.

% Systems exist already for the conversion of piano keyboard. The Yamaha
% disklavier, the Don Buchla designed MOOG Piano Bar, Bosendorfer, PNOscan
% by QRS focus on reproduction. These same systems could be applied to a
% harpsichord, but the limitation is a single data stream per key.

% \subsection{Motivations}\label{motivations}

% \begin{itemize}
% \item
%   existing sysetms
%   \begin{itemize}
%   \item
%     Yamaha Disklavier used as a basis for exploration into haptics and
%     the piano keyabord
%     \cite{MusicalHaptics2018_04, MusicalHaptics2018_05,
%     MusicalHaptics2018_13}
%   \item
%     Bosendorfer: LED photo transistor pair \cite{Moog1990} for obtaining
%     performance data \cite{MusicalHaptics2018_05}

%     \begin{itemize}
%     \item
%       290SE
%     \item
%       CEUS
%     \end{itemize}
%   \item
%     Don Buchla / Moog Piano Bar
%   \item
%     PNOscan MIDI 9 QRS Music as MIDI Piano system \cite{McPherson2013}

%     \begin{itemize}
%     \item
%       PNOmation
%     \end{itemize}
%   \item
%     Piano Disc
%   \end{itemize}
% \end{itemize}

@@

% These systems mentioned could be adapted for a harpsichord, but they would not
% easily capture the separate plucks from both jacks per key. We decided to create
% our own system and focus on detecting the vertical displacement of the jacks
% directly, rather than the position of the key.

% Methodologies for increasing visitor engagmemnet or

% Museum staff want visitors to be engaged but limits in staffing and
% funding limit how vistors can interact with collections
% \cite{Templeton2018,
% McAlpine2014}. For musical instrument museums there is an additional
% difficulty in the balance of keeping historical instruments playable
% \cite{McAlpine2014}. The fragility and decay of musical instruments
% means that there is an inveitable point where instruments will no longer
% be in a playable condition \cite{McAlpine2014, Fritz2017}

% McAlpine outlines a similar situation to the Tagliavini collection in
% his work with the Benton Fletcher Collection \cite{McAlpine2014}. There
% was a stipulation by Benton Fletcher that the instruments remain
% available to play when they were gifted to National Trust Fenton House.
% A custom MIDI interface arranged in a two manual harpsichord style which
% triggered samples of the instruments in the collection using an
% appropraite recording strategy for each. One problem highlighted in user
% tests by McAlpine is that the weighted keys did no provide ``an
% experiential sense of interacting with a historical keyboard''
% \cite{McAlpine2014}. The MIDI interface created from commercially
% available weighted keys. McAlpine posits a haptic keyboard augemented
% with ``actuators to provide positionally-sensitive real-time force
% feedback at point-of-contact'' in a similar approach as the piano
% mechanism by Gillespie \cite{Gillespie1996}.

% Previous NIME project ``Tromba Moderna'' \cite{Baldwin2016} looked to
% avoid the complex engineering problem by simply recreating the tromba
% marina and augmenting it. In the case of the Tromba Modern, a piezo
% transducer was connected to a sound synthesis engine connected to a
% driver inside the instrument to simulate the vibration that would be
% expected of an historical tromba marina.

% The original design was to take a similar approach to the project by
% Timmermans \cite{Timmermans2020}, which was an extension of the system
% by Gillespie \cite{Gillespie1996}. The process laid out in Timmermans
% project was to begin with a single key model of the mechanism. We began
% with a 3-key Model Harpsichord Mechanism by Graziano Bandini
% \textbackslash figure\{\}. Before we continued with augementing a model
% with actuators it was decided that it we should first validate whether a
% recreation of the harpsichord mechanism would suffice in allowing
% visitors to suspend disbelief.

% Benefits of putting effort into the maintenance of one interface with a
% transferable sensor system

% An additonal factor not faced in the Tromba Moderna project is the
% problem of scale in managing 98 rather than a single data stream of
% data.

% Material costs mitigated by creating a scaled version of the sound
% board, providing only the length of string necessary under comparable
% tension and thus resistance to the jack quill.

% Finally, a further extension to the projects above was a commmitement to
% open sourcing all aspects of the project, hardware, software, and data.

% There are commercial systems for augmenting piano keyboards generating MIDI data.

% An additional goal of the project was to design an interface that could be used to research
% the influence the mechanism and it's resistances have on a player's performance.
% For example, the Yamaha Disklavier series of pianos have an embedded sensor
% system that allows for the instruments to operate in a "silent mode", but still
% output MIDI data when the keys are struck. This has been a rich avenue for
% exploring the affect of vibration on performance,
% \cite{MusicalHaptics2018_04,MusicalHaptics2018_13} perception in instrument
% quality \cite{MusicalHaptics2018_05}.

% As well as a trying to offer a more engaging experience to museum attendees, the
% interface was also designed as a potential probe for further research into the
% idiosyncrasies of the harpsichord and the influence the mechanism has on
% performance.

Museums often face significant challenges in engaging visitors due to limitations in staffing and funding, which restrict the ways in which visitors can interact with collections \cite{Templeton2018, McAlpine2014}. For musical instrument museums, these challenges are compounded by the difficulty of preserving historical instruments in a playable condition \cite{McAlpine2014}. The instruments' inherent fragility and gradual decay inevitably result in a point where they can no longer be played, even when collections adhere to the strictest conservation protocols. A significant cultural change has taken place in recent decades, shifting the focus from the playability of the originals to their conservation. Karp \cite{Karp1979,Karp1985} advocates for a deeper understanding of musical instruments so that enough knowledge is generated to make them as ``copyable'' as possible.

McAlpine discusses a case similar to the Tagliavini collection in his examination of the Benton Fletcher Collection at National Trust Fenton House \cite{McAlpine2014}. When these instruments were donated, Benton Fletcher stipulated that they remain playable and should continue to be maintained for tuition and public performance. A large sampling campaign was conducted, and a custom MIDI interface was designed to fulfil this requirement while preserving the original instruments' integrity. The MIDI keyboard, comprising two commercially available keyboards mimicking the two-manual harpsichord layout, were used by visitors to trigger the instrument samples recorded with tailored strategies for each. However, user tests identified a significant limitation: the commercially available weighted keys failed to provide an authentic sense of interacting with a historical keyboard \cite{McAlpine2014}.

% To address this, McAlpine suggested designing a haptic keyboard enhanced with actuators capable of delivering positionally sensitive, real-time force feedback at the point of contact, inspired by previous projects such as Gillespie’s work on piano mechanisms \cite{Gillespie1996}.

@@

\section{Design}\label{design}

The present project initially drew inspiration from Timmermans' extension of Gillespie’s system \cite{Timmermans2020, Gillespie1996}. Timmermans’ process began with a single-key model of the mechanism. Similarly, we started with a 3-key model harpsichord mechanism, crafted by Graziano Bandini. Before incorporating actuators into the model, we sought to validate whether a faithful recreation of the harpsichord mechanism could effectively enable visitors to suspend disbelief.

\begin{figure}  
  \centering
  \includegraphics[width=\linewidth]{src/images/3-key-side-labelled.png} 
  \caption{3-key Model Harpsichord Mechanism by Graziano Bandini, Bologna 2023. 3-Key modelled used for initial prototyping. A: Wrestplank B: Key Lever C: Key cap D: soundboard E: Jack} 
  \Description{A 3-Key Model of a Harpsichord Mechanism created Graziano Bandini in Bologna. A side view of the model which is a cross-section of what is a common
mechanism for an Italian style harpsichord of the XVI century.} 
  \label{fig:3key}
\end{figure}

A critical benefit of focusing on the development of a single interface with a transferable sensor system is the potential for scalability and adaptability. However, managing data streams presents a significant challenge. Unlike the Tromba Moderna project, which dealt with a single data stream, our design must manage 98 simultaneous data streams. This complexity necessitated cost-effective solutions, such as scaling down the soundboard to include only the necessary string length under comparable tension and resistance to the jack quill.

Another distinguishing feature of this project is its commitment to open-sourcing all hardware, software, and data components. While commercial systems for augmenting piano keyboards and generating MIDI data exist, this project emphasizes accessibility and transparency.

An additional objective was to design an interface that supports research into how the harpsichord mechanism and its resistances influence a player’s performance. For example, the Yamaha Disklavier series incorporates embedded sensors enabling "silent mode" operation while outputting MIDI data when keys are struck. This has facilitated research into the effects of vibration on performance \cite{MusicalHaptics2018_04, MusicalHaptics2018_13} and the perception of instrument quality \cite{MusicalHaptics2018_05}.

\section{Design}

The design criteria was to create a keyboard interface that provided a comparable tactile response to that of a harpsichord in the italian style.

Accommodations were made in the design to allow for as much flexibility as possible. 
This also meant that designs for the electronics would have to include a level of flexibility in how they might be installed. 
Surplus space was added into the design of the internal chamber of the model Figure \ref{fig:49-key-bottom}, but there was still a limit given the structural
and aesthetic requirements.

\section{Constraints}\label{design-problem}

% Central philosophy of the Tagliavini Collection is that the instrument
% should remain playable to the public. The condition of some instruments
% is such that restoration would required a replacement of vast majority
% of the instrument, would be too costly, would risk the instrument given
% its fragility, or all of those in combination.

% This is not an uncommon problem faced by musical instrument museums
% worldwide. The central question the project tries to answer is ``how do
% we conserve musical instruments in order to continue interacting with
% them''

% The approach taken was to recreate the interface of these instruments
% not for acoustics purposes be purely for the mechanical response. As
% exhibition for the general public in a musical instruments there were
% aesthetic and functional constraints placed on the design of the
% interface.

% \subsection{Design Constraints}\label{design-constraints}

There were constraints that were put in place by the Museum on the design of the exhibit.

\begin{itemize}
\item
  No electronics should be visible \footnote{with the exception of a pair of headphones for listening and a display to change parameters of the audio engine}
\item
  It should be robust and reliable, easily maintained by museum staff without complex technical processes
\item
  it should not augment or fix the mechanics, but present them with their original limitation
\end{itemize}

% Another constraint was that the design of the sensor system and the
% model keyboard would have to carried out at a distance in separate
% workshops. Time constraints meant the keyboard was being fabricated for
% a sensor that did not yet exist.

\subsection{Building the Model Keyboard}\label{keyboard}

% PARAGRAPH FROM ROBERTO
The construction of the keyboard required three months of work to be fully completed
and followed the same construction phases of a real harpsichord.

The model was constructed using the materials of the Italian harpsichord
tradition. Parts are labelled in Figure \ref{fig:3key}.

\begin{itemize}
\item
  Walnut wood for the wrestplank
\item
  Chestnut wood for the key levers
\item
  Boxwood and ebony for the key caps.
\item
  Cypress wood for the soundboard
\item
  Beech wood, a brass spring and a seagull feather plectrum for the 98 jacks
\end{itemize}

The body of the case is based on a rectangular poplar frame that leaves space
for the installation of the electronic sensors.

Deviations from the design of the Trasuntino are the rectangular, rather than
'logarithmic' shape, removal of a bottom panel and omission of a short octave,
instead spanning 4 octaves beginning on C.

The strings are Malcolm Rose yellow brass wire, which are secured to wrought
iron pins. The tension of each string was adjusted to provide resistance to
plucking close to that of a harpsichord. Three felt strips were woven between
the strings in order to dampen the vibration of the strings.

The result is an interface instrument that is placed between the mechanical
action of the keys and the generation of a synthetic sound. The touch of the
wooden keys, the friction of the natural feathers on the string, the noise of
the percussion at the base of the key together produce the sensation of being in
front of a real harpsichord.

% \subsubsection{Anatomy of Harpsichord
% mechanics}\label{anatomy-of-harpsichord-mechanics}

% \begin{itemize}
% \item
%   simple
% \item
%   aesthetics
% \item
%   split workload
% \item
%   parts of harpsichord

%   \begin{itemize}
%   \item
%     jack

%     \begin{itemize}
%     \item
%       quill
%     \item
%       jack body
%     \item
%       pivot
%     \item
%       tongue
%     \end{itemize}
%   \item
%     key
%   \item
%     body
%   \end{itemize}
% \end{itemize}

% The decision was made to design the electronics and the model keyboard simultaneously. 
% This meant development development time was quicker and iteration on the idea.
% Likely the first implementation would need re-designed and redesiginng
% the electronics to fit the form of the keyboard was more cost effective
% than building or carrying out major modifications to the model.

% The model was designed to have two internal chambers a front and rear.
% The front and back that was xxx by yyy by 370mm between the faceplate
% and the jacks. The TONGUE THAT GOES INTO SLOT divides the chamber through
% a small gap between the underside of the soundboard and the key guides.
% The key guide needs to be deep enough to allow for the full movement of
% the key and stop the key becoming unseated. The chamber then extends to
% a support wall which divides the front and rear.

% Discussion of spatial constraints and effects on electronics design and
% installation is covered in \textbackslash section\{setup\}

% The rear chamber was designed to house components for audio processing,
% which is discussed further in \textbackslash section

\begin{figure}  
  \centering
  \includegraphics[width=\linewidth]{src/images/49-key-bottom-sensors-no-keys.jpg} 
  \caption{Underside of the full model keyboard showing two chambers. Front chamber at the top and rear chamber at the bottom.} 
  \Description{} 
  \label{fig:49-key-bottom}
\end{figure}

% \footnote{https://github.com/Nemus-Project/haptic-harpsichord-nime-2025/issues/Nemus-Project/haptic-harpsichord-nime-2025/images/disegno.pdf}

% A harpsichord that
% was economic enough to be deconstructed would also have been of a more
% modern mechanical design and not of the design that was being recreated
% with the model. Prototyping stage would likely be invasive and damage or
% heavy modification likely. As such it was not a good use of funds as a
% modern instrument fit for destruction did not offer benefits that the
% small model with easy internal access did not already provide.

% \subsubsection{3-Key Prototype}\label{key-prototype}

% \begin{itemize}
% \item
%   proof of concept
% \item
%   available on loan from museum
% \item
%   easy access to internal structure
% \end{itemize}

% \subsubsection{49-Key Prototype}\label{key-prototype-1}

% For the final interface a 49-key (4-octave) was constructed by luthier
% \anon{Roberto Livi}. The keyboard is in the same style† as the 3-key model with
% to jacks per key.

% The keyboard is fully enclosed and internally has two chambers, front
% and rear, in which the electronics could be installed
% \textbackslash figure(underside 49-key).

% Keyboard to be used for exhibition at san colombano where it would be
% available to play by general public.

% Timescale: from initial discussion to delivery of interface to the \anon{NEMUS}
% lab was 8-months. Active assembly of the instrument was 3 months

% Constructed From:

% \begin{itemize}
% \item
%   for exhibition
% \item
%   problems of scale
% \item
%   problems of timescale
% \end{itemize}

@@

\begin{itemize}
\item
  Non-invasive: The sensor system should not require major modifications
  to the interface on which it is installed.
\item
  Low Latency: Reading sensors and then triggering an output should have
  a low latency. Following \cite{Jack2016} an latency was set as 10 ms
  with an upper limit set to 25ms beyond which the sensors would be
  deemed unusable
\item
  Reliable: Data obtained should be reliable and accurate enough not to
  result in false positives or false negatives without extensive
  filtering
\item
  Scalable: The system should be scalable both economically and
  temporally. Given the open source nature of the project, it could not
  be prohibitively expensive to carry out. Also any system that would
  work on a single key would have to scale financially to potentially 50
  or more. The time required to install and calibrate the sensors would
  also have to scale.
\item
  Expandable: Any sensor system should be functionally expandable to
  allow for the usage of the data for control over other parameters.
\end{itemize}

% \subsection{Prototyping}\label{prototyping}

% \subsubsection{Triggering a note}\label{triggering-a-note}

% The message format covered in \textbackslash section

% \section{Related Work and Motivation}\label{related-work-and-motivation}

% \begin{itemize}
% \item
%   \cite{Timmermans2020}
% \item
%   \cite{McPherson2013} \cite{McPherson2019}
% \item
%   \cite{Mudd2013}
% \item
%   \cite{Fritz2017}
% \end{itemize}

% \section{Hardware Design}\label{hardware-design}

% The hardware was designed with the xx in mind.

% Using a similar apporach to @HapticKey the first step was to fabricate
% or procure a model of a single working key. San Colombano provided a
% model 3-key of a harposchord mechanism by Graziano Bandini
% \textbackslash FIGURE. The 3-key model was used a test-bed for potential
% sensors while designs were drawn for a full-scale model. The
% cross-sectional nature of the model meant it was easy to fit and remove
% sensors. Small size meant that it would also be easy to transfer between
% workshops.

% 3-key model also had a sliding jack rail which allowed for easy
% disengagement of the physical mechanical response, which allowed for
% some preliminary tests on the presence and absence of mechanism.

% \subsection{Failed sensors}\label{failed-sensors}

% A number of sensors were tried before the final system was decided upon.
% An inertial measurement unit (IMU) was mounted to a single key which
% contained a 6-degrees-of-freedom accelerometer and gyroscope. There was
% a consistent set of conditions to trigger sound, but this was not easily
% differentiable when other keys were played. The fitting time and cost of
% a single IMU could b reduced when purchased in bulk but was far greater
% than other systems considered.

% A reed sensor and a hall sensor were mounted and magnets embedded within
% the jacks. Adjusting the height of the sensor provided a reliable means
% of detecting a threshold, in particular the hall sensors which could be
% partially calibrated physically and also in software. The time taken to
% calibrate a single key was time consuming. Ideally each sensor would be
% placed on a thread for adjustment, but given the constraints on the
% internal space it was decided any adjustment of heights would likely
% mean an unacceptable length of time to install and calibrate. Embedding
% magnets into the jacks was also did not satisfy the criteria of being
% scalable and non-invasive.

% Light dependant resistor mounted to the jack rail. as jacks were pressed
% less light would be let through. Usable data but very dependant on
% lighting conditions. Modifying the jack also did not satisfy the
% non-intrusive criteria and the placement outside the internal chamber of
% the model meant hiding electronics would be more difficult.

% Finally, force sensitive resistors placed under the jack of each key was
% considered but was never implemented. The sensors were to be placed
% under each jack either above or below the padding on which the bottom of
% the jack rests.

% Further discussion of this approach in \textbackslash section.

% \section{Electronics}\label{electronics}

% Design of electronics carried out in tandem with construction of full
% model. Since the Electronics were designed at a distanced to the final
% full model, some flexibility was required in the design. Design decisions
% that were not as optimal but would allow for change in wiring and
% fitting should there be any issues during installation.

@@

% The equipment required including soldering suitable for through-hole technology (THT) components and 3D printing.

% Laser cutting and CNC routing facilities were used during the research
% for the project but not required for the final implementation.

@@

% NEMUS lab under construction during the first section of the project.
% Facilities available at the maker spaces of \anon{Bologna (Alma Labor)} and
% \anon{Edinburgh (uCreate)} used for what was deemed reasonable assumption in
% terms of available equipment.

@@

Another common application of the QRE1113 is within line following
robots, where an array of QRE1113s are used in conjunction with black
tape or reflective. A motorised system can then use this data to follow the line of tape and adjust its position.

Applied to the jacks was a greyscale gradient printed the length full
travel range of the jack.

For the first iteration this was performed on an inkjet printer and
double-sided tape. For the final iteration the stickers were printed on
(Coala, 1D 100 Gloss P, gloss, white, permanent adhesive, 300g/m2, 100
µm, 1370mm x 50.00m, Solvent/Latex/UV) using a HP Latex 115 and then
cutout with a Summa 150 roll cutter. The printer and roll cutter greatly
reduced the manual labour for cutting stickers meaning the process would
scale from the 3-key to the 49-key model.

% An initial sticker design was used that contained only the gradient. It
% was found that the jacks were sticking in their slots and the edges on the
% stickers easily peeled around the printing The sticker was redesigned to
% be the full length of the jack which avoided the problem of peeling. The
% larger are also mean the stickers adhered better to the jack Some
% trimming was still required of each sticker to avoid them adding extra
% resistance to the jack's movement.

During testing it was found that their was a considerable amount
cross-talk between adjacent sensors. This meant the reading from key
would be affected by the movement of the key next to it. The cross-talk
was assumed to be as a result of light reflecting from other jacks.
Given that IR is used this was difficult to confirm by sight. A set of
baffles was designed for the PCB that would wrap round the jack and stop
light from other jacks or sensors. The baffles were printed on a BAMBU X1 3D
printers using PLA plastic. A dark pigmented PLA was used to avoid IR
light simply being diffused. After the baffles were fitted, the
cross-talk was eliminated.

\subsubsection{RGB LEDs}\label{rgb-leds}

for problems of scale:

Addressing the problem of scale of identifying which key's threshold is
currently being edited.

The LEDs also allowed for easy identification for keys that were outside
of expected values and when a key was triggered. This allowed for visual
identification of malfunctioning or un-calibrated keys.

The LEDs are on the same PCB as the sensors and are hidden when the
name board is in place.2024

Would suggest that this is a necessity of this kind to provide a means
to identify and diagnose LEDs with state.

TLC5940 or other PWM driver such as in \cite{McPherson2013} added
complexity to assembling, but were avoid in favour of Addressable LEDs
with integrated driver. The integrated LEDs share the same data line
which has the limitation where a faulty LED will stop all following LEDs
from working. An individual PWM driver per PCB would allow avoid
situations where failure cascades, but would add complexity in assembly
as the parts are typically SMD. Also, given the LEDs are fixed to the
board and could not be repaired in situ, the solution would be to
replace the PCB regardless of which approach was taken. Given the LEDs
were only used in debugging and calibration and otherwise would not be
visible, a simpler design approach was taken.

\subsubsection{Multiplexing}\label{multiplexing}

The BLE Nano has 8 ADC pins connected to one ADC channel via a multiplexer. 
We avoided the complexity of communicating multiple microcontrollers and instead used a Texas Instruments CD4051B multiplexer on each sensor board.
An 8-channel multiplexer was chosen as the model has 49-keys and a division of 7 PCBs each containing 7 sensors.
The multiplexer reduced the 7 sensor signals to a single channel. The CD4051B is addressed with a 3-bit which necessitates 3 digital channels.
Transistor / Op-amp circuit could have been used similar to the one in
\cite{McPherson2013}. Given time constraints it was beneficial to continue using and entirely THT based designed for easy assembly.

% \subsection{PCB design}\label{pcb-design}

% EAGLE PCBs designed using EAGLE.

% Worth noting that AutoCAD has
% deprecated EAGLE and the programme will no longer be supported from
% 2026 Project files will need to be transitioned to another standard,
% likely the KiCAD programme.

% trade off between limiting number of boards required and allowing
% compensation for variation in pitch.

% Smaller PCBs means that spacing could be adjusted to accomodate the
% changes in pitch between jacks. It also allowed for a modular approach
% where any PCB could swapped for minimal cost.

>>>>

@@

Museums often face significant challenges in engaging visitors due to limitations in staffing and funding, which restrict how visitors can interact with collections \cite{Templeton2018, McAlpine2014}. For musical instrument museums, these challenges are compounded by the difficulty of preserving historical instruments in a playable condition \cite{McAlpine2014}. The instruments' inherent fragility and gradual decay inevitably result in a point where they can no longer be played, even when collections adhere to the strictest conservation protocols \cite{NYT_strad}. A significant cultural change has taken place in recent decades, shifting the focus from the playability of the originals to their conservation. Karp \cite{Karp1979,Karp1985} advocates for a deeper understanding of musical instruments so that enough knowledge is generated to make them as ``copyable'' as possible.

McAlpine discusses a case similar to the Tagliavini collection in his examination of the Benton Fletcher Collection at National Trust Fenton House \cite{McAlpine2014}. When these instruments were donated, Benton Fletcher stipulated that they remain playable and should continue to be maintained for tuition and public performance. A large sampling campaign was conducted, and a custom MIDI interface was designed to fulfil this requirement while preserving the original instruments' integrity. The MIDI keyboard, comprising two commercially available keyboards mimicking the two-manual harpsichord layout, was used by visitors to trigger the instrument samples recorded with tailored strategies for each. However, user tests identified a significant limitation: the commercially available weighted keys failed to provide an authentic sense of interacting with a historical keyboard \cite{McAlpine2014}.

@@

\section{Design Principles}\label{design}

\begin{figure*}
\includegraphics[width=\linewidth]{src/images/details.jpg}
\caption{Details of the replica, including the key slots and sides, keys, and jacks with plectra.}\label{fig:details}
\end{figure*}

The design of the augmented replica keyboard for the Tagliavini Collection was guided by three principal constraints stipulated by the museum:
\begin{itemize}
\item \emph{No visible electronics}. To preserve the visual integrity of the exhibit, all electronic components had to remain concealed except for a pair of headphones and a small display for audio parameter adjustments.
\item \emph{Robustness and reliability}. The system needed to accommodate frequent use by museum visitors and allow for straightforward maintenance by staff without requiring specialised technical expertise.
\item \emph{Faithfulness}. The keyboard mechanism ensured fidelity to its authentic operation, keeping artefacts to a minimum. 
\end{itemize}
These constraints demanded a non-invasive, robust design that was respectful of the historical authenticity of the instruments. The requirement for faithfulness to the original mechanism immediately ruled out a mechanical design reliant on electromechanical actuators, as in previous works on piano haptics \cite{Timmermans2020,Gillespie1996}. While actuators can generate considerable force, they cannot replicate the wideband signals characteristic of the harpsichord's sharp and responsive haptics (A REFERENCE WOULD BE GOOD HERE, ANY IDEAS?), which differ significantly from those of a piano.

\subsection{Summary of the Finalised Design}

The final design is a 49-key harpsichord replica equipped with an optical sensor system to reproduce the tactile characteristics of the original instrument, visible in Figure \ref{fig:teaser}. Several key features were implemented to address the project’s requirements:

\begin{itemize}
\item \emph{QRE1113 infrared sensors}. These sensors were selected for their non-invasive properties and precision, enabling the detection of jack displacement without requiring structural modifications to the keyboard mechanism.
\item \emph{Gradient stickers}. Greyscale gradient stickers were applied to each jack to allow the infrared sensors to make precise readings. This approach made calibration easier and enabled scalability from a 3-key prototype to the final 49-key design.
\item \emph{Jack baffles}. Custom-designed baffles, 3D printed from dark pigmented PLA to minimise infrared reflection, were installed around the jacks to prevent cross-talk between adjacent sensors and ensure reliable data capture.
\end{itemize}

\begin{figure}  
  \centering
  \includegraphics[width=\linewidth]{src/images/49-key-bottom-sensors-no-keys.jpg} 
  \caption{Underside of the full model keyboard, showing two chambers: the front chamber (top) and the rear chamber (bottom).} 
  \Description{} 
  \label{fig:49-key-bottom}
\end{figure}

A modular system of PCBs was designed to manage the sensors and process their output. The 49 QRE1113 sensors were distributed across seven boards, containing 7 sensors each. The PCBs were secured to the underside of the keyboard mechanism, allowing them to be adjusted during installation. Ribbon cables connected the PCBs, providing flexibility during assembly while maintaining a compact form factor. Additional modifications, including baffles and adhesive improvements, were made to optimise the reliability of the sensor system during calibration and use.

The project built upon earlier NIME research on generating MIDI messages from piano keystrokes \cite{McPherson2013}, adapting it to address the unique characteristics of harpsichords and the specified constraints. Whereas the previous design emphasised continuous gesture tracking, this implementation required discrete key-triggered data to align with the needs of MIDI-triggered audio playback. The adoption of gradient stickers in place of continuous optical tracking embodies this adaptation, striking a balance between precision, cost-efficiency, and practicality.

\subsection{Materials and Construction}

The keyboard was designed to replicate the tactile and aesthetic sensations of playing an antique Italian harpsichord. Traditional materials were used, including walnut for the wrestplank, chestnut for the key levers, boxwood and ebony for key covers, and cypress for the case and soundboard. The 98 jacks were made from beech, fitted with brass springs and natural seagull feather plectra. The design was inspired by Venetian harpsichords, particularly the Alessandro Trasuntino instrument at the San Colombano Museum. Details are visible in Figures \ref{fig:teaser} and \ref{fig:details}.

The rectangular poplar frame deviates from the traditional logarithmic form to allow for the installation of the electronic sensors, as visible in Figure \ref{fig:49-key-bottom}, without compromising the visual or tactile authenticity. Two string orders, crafted from yellow brass wire and anchored with wrought iron pins, were tensioned to replicate authentic plucking resistance. Felt strips were added to dampen vibrations.

The result is an interface that combines the mechanical action of keys with synthetic sound generation, preserving a real harpsichord's tactile and auditory qualities.

@@

\begin{itemize}
\item
  Non-invasive: The sensor system should not require major modifications
  to the interface on which it is installed.
\item
  Low Latency: Reading sensors and then triggering an output should have
  a low latency. Following \cite{Jack2016} an latency was set as 10 ms
  with an upper limit set to 25 ms beyond which the sensors would be
  deemed unusable
\item
  Reliable: Data obtained should be reliable and accurate enough not to
  result in false positives or false negatives without extensive
  filtering
\item
  Scalable: The system should be scalable both economically and
  temporally. Given the open source nature of the project, it could not
  be prohibitively expensive to carry out. Also any system that would
  work on a single key would have to scale financially to potentially 50
  or more. The time required to install and calibrate the sensors would
  also have to scale.
\item
  Expandable: Any sensor system should be functionally expandable to
  allow for the usage of the data for control over other parameters.
\end{itemize}

@@

@@

@@

The optical sensors were directed at a greyscale gradient printed on a sticker and applied to the side of the jack.
This was first trialled using a generic inkjet printer and double-sided tape. 
For the final iteration, the stickers were printed on Coala 1D 100 Gloss PG 100 $\mu$m white gloss monomeric PVC with permanent grey acrylic dispersion adhesive \footnote{Datasheet: \url{https://www.antalis.co.uk/mediashare/g4media/pdf/TS_EN_COALA_1D_100_GLOSS_P_00_ISS_15022023.pdf}} using a HP Latex 115 and then cut with a Summa 150 roll cutter. 
The printer and roll cutter greatly reduced the manual labour for cutting stickers meaning the process would scale from the 6 jacks of the 3-key to the 98 jacks of the 49-key model.

\begin{figure}  
  \centering
  \includegraphics[angle=90,width=\linewidth]{src/images/jack-w-tag.jpeg} 
  \caption{} 
  \Description{} 
  \label{fig:jack-w-tag}
\end{figure}

During testing it was found that their was a considerable amount cross-talk between adjacent sensors. 
This meant the reading from key would be affected by the movement of the key next to it. 
The cross-talk was assumed to be as a result of light reflecting from other jacks.
Given that IR is used this was difficult to confirm by sight. 
A set of baffles was designed for the PCB that would wrap round the jack and stop light from other jacks or sensors \ref{fig:baffles}. 
The baffles were printed on a BAMBU X1 3D printers using PLA plastic. A dark pigmented PLA was used to avoid IR light simply being diffused. 
After the baffles were fitted, the cross-talk was eliminated.

\begin{figure}  
  \centering
  \includegraphics[,width=\linewidth]{src/images/baffles.png} 
  \caption{} 
  \Description{} 
  \label{fig:baffles}
\end{figure}

Calibration to pluck thresholds were to be carried out manually.  To allow for easy identification of the sensor being addressed during the calibration process, RGB LEDs were integrated into the design. The LEDs were also used in the identification of any malfunctioning or un-calibrated sensor.

TLC5940 or other PWM driver such as in \cite{McPherson2013} add robustness should an LED fail, but complexity to assembly process. 
Instead, the design uses addressable LEDs with integrated driver that share the same data line. 
This has the limitation where a faulty LED will stop all subsequent LEDs in the line from working.

The BLE Nano has 8 ADC pins connected to one ADC channel via a multiplexer. 
We avoided the complexity of communicating multiple microcontrollers and instead used a Texas Instruments CD4051B multiplexer on each sensor board.
An 8-channel multiplexer was chosen as the model has 49-keys and a division of 7 PCBs each containing 7 sensors.
The multiplexer reduced the 7 sensor signals to a single channel. The CD4051B is addressed with a 3-bit which necessitates 3 digital channels.
Transistor / Op-amp circuit could have been used similar to the one in \cite{McPherson2013}.
